The ppmData R-package for setting up spatial point process models

### 2022-08-10 paper/paper.Rmd
a few more edits

## Generate a quadrature scheme using quasi-random sampling

In the `ppmData` package we have tried to make the generation of a quadrature scheme of a single or marked point process as easy as possible. We provide a simple interface for generating a Poisson point process data object and base all inputs and outputs using `terra` [@hijmans_terra_2021] and `sf` [@pebesma_simple_2018] packages. All we need to pass is a set of coordinates from a point process and we can generate a very basic point process data object. Ideally, the user knows the region they are interested in modelling and this is passed as a window, we use a `terra` SpatRaster object. The window defines the region where we expect to encounter observations and where to generate the quadrature locations. There are a number of other calls which can be passed to `ppmData`, but the main one is the inclusion of covariates which can be used in an Inhomogenenous Poisson Process Model. Under this scenario pass a set of covariates that share the same extent and resolution as the window SpatRaster. The values of the covariates will be extracted at the locations of the quadrature scheme. Whilst not unique to this package, the `ppmData` package can also generate grid and pseudo-random quadrature schemes as presented in Fig. 1.

@@

One of the main reasons for creating the `ppmData` package was to develop quadrature schemes for multiple species (marked) point process. The multiple species quadrature scheme can be used in multiple species models available for use in the `ecomix` package [@woolley_ecomix_2022], or models like joint species distribution models [@ovaskainen_making_2011]. Here is a quick example of how we would set up a quasi-random quadrature scheme for multiple species. The main difference is the the column `speciesID` contains multiple species, `ppmData` recognises this internally and subsequently sets up a marked point process quadrature scheme. The quadrature scheme has a common set of quadrature locations, but the quadrature weights are calculated on a species specific basis $w_{ij}$, where $i$ is a sites index and $j$ is a species index.

@@

For the purpose of demonstration, we show how to predict the model using `terra` and a `glm` object. This will return the expected intensity when all the areal weights of each cell in the raster equal to one. To get the intensity of *Tasmaphena sinclairi* per cell re-scale the intensity based on the area represented by each raster cells. This essentially gives us the expected count of points (species presences) per raster cell. See Figure 3 for an example output.

>>>>

## Generating a quasi-random quadrature scheme

In the `ppmData` package we have tried to make the generation of a quadrature scheme of a single or marked point process as easy as possible. We provide a simple interface for generating a Poisson point process data object and base all inputs and outputs using `terra` [@hijmans_terra_2021] and `sf` [@pebesma_simple_2018] packages. Users need to pass a set of coordinates from an observed point pattern, a window to define the point pattern region and a set of covariates observed across this region. The window is defined using a `terra` SpatRaster object, and it used to identify the point pattern region and where to generate the quadrature locations. Covariates are included as a multiple layered `terra` SpatRaster object that share the same extent and resolution as the window. The values of the covariates will be extracted at the locations of the point pattern and quadrature scheme. Whilst not unique to this package, the `ppmData` package can also generate grid and pseudo-random quadrature schemes as presented in Fig. 1.

@@

One of the main reasons for creating the `ppmData` package was to develop quadrature schemes for multiple species (marked) point process. The multiple species quadrature scheme can be used in multiple species models available for use in the `ecomix` package [@woolley_ecomix_2022], or models like joint species distribution models [@ovaskainen_making_2011]. Here is a quick example of how we would set up a quasi-random quadrature scheme for multiple species (Fig. 2). The main difference is the the column `speciesID` contains multiple species, `ppmData` recognises this internally and subsequently sets up a marked point process quadrature scheme. The quadrature scheme has a common set of quadrature locations, but the quadrature weights are calculated on a species specific basis $w_{ij}$, where $i$ is a sites index and $j$ is a species index.

@@

Finally, we show how to predict the model using `terra` and a `glm` object. This will return the expected intensity when all raster cell areas equal to one. To get the intensity of *Tasmaphena sinclairi* per cell, we need to re-scale the intensity based on the area represented by each raster cells. The re-scaled intensity returns the expected count of points (species presences) per raster cell. See Fig. 3 for an example output.
